An investigation into the contentious phylogenetic results of the *Mixta* species

### 2021-05-03 Manuscript/Manuscript_Paper.Rmd
Update Manuscript_Paper.Rmd

<!-- \newpage -->
# Abstract
<!-- *Mixta* is a recently described genus within the bacterial family *Erwiniaceae* and is closely related to *Pantoea*, *Erwinia*, and *Tatumella*. Depending on the genes and method used for phylogenetic analysis, the genus with which *Mixta* shares the most recent ancestor differs. This study aimed to determine the cause behind these contentious results and identified genes that may have resulted from horizontal gene transfer. Ten complete or partial genomes – two *Mixta* species, six species from other *Erwiniaceae* genera, and two outgroups – were retrieved from NCBI and annotated. Homologous genes were extracted yielding a dataset of 799 genes. The genes were aligned with the ClustalW algorithm, and MEGA-CC was used for model testing and computing phylogenetic trees for each gene for both nucleotide and amino acid sequences. Distance matrices were then extracted from these trees. Nucleotide and amino acid identity analyses were also done using the programming language `R`. *Pantoea* was the closest relative to the *Mixta* species in most analyses; however, results were not consistent. Some genes were also found to be more similar to other, non-*Pantoea* species. Diligence must be given to selecting genes for phylogenetic analysis and the method chosen to prevent any xenologous signal from distorting the actual relationships. Furthermore, future research should consider that different phylogenetic analyses may provide different results. -->

# Intro and MM
<!-- \newpage -->
<!-- # Introduction -->
<!-- The bacterial family *Erwiniaceae* includes diverse Gram-negative [@Winslow1917; @Prakash2015] organisms, including the genera *Mixta*, *Pantoea*, *Erwinia*, *Tatumella*, *Buchnera*, *Wigglesworthia*, *Phaseolibacter*, and *Kalamiella* [REF CRAIG]. *Mixta* is a relatively newer genus first identified by Palmer et al. in 2017 [@Palmer2017] and later described by the same researchers in 2018 [@Palmer2018]. Currently, there are only five *Mixta* species validly published [@LPSN]. *M. calida* and *M. gaviniae* (previously species of *Pantoea*) were initially found in infant formula and its production environment [@Popp2010]. Fritz et al. [@Fritz2014] demonstrated that some strains of *M. calida* were linked with post-surgical meningitis in humans. The other *Mixta* species include *M. theicola* which was isolated from black tea extract [@Kato2015], *M. intestinalis* which was isolated from a healthy human fecal sample [@Prakash2015], and *M. tenebrionis* which was isolated from the gut microbiota of a plastic-eating mealworm [@Xia2020]. The first two were previously named as *Pantoea* species. -->

<!-- *Mixta*'s position relative to other members of the *Erwiniaceae*, in particular to *Pantoea*, *Erwinia*, *Tatumella*, differs vastly between studies [@Prakash2015; @Palmer2017; @Palmer2018; @Adeolu2016; @Brady2011; @Rezzonico2016; @Brady2012; @Gueule2015]. Prakash et al. [@Prakash2015], Palmer et al. [@Palmer2017], Rezzonico et al. [@Rezzonico2016], and Gueule et al. [@Gueule2015] all used *gyrB*, *rpoB*, *atpD*, and *infB* nucleotide gene sequences in their phylogenetic analyses. Prakash et al. [@Prakash2015] conducted a neighbour-joining statistical method while the others used a maximum-likelihood (ML) method [@Palmer2017; @Rezzonico2016; @Gueule2015]. Three of the studies – Rezzonico et al. [@Rezzonico2016], Prakash et al. [@Prakash2015], and Gueule et al. [@Gueule2015] – placed the *Mixta* species within the *Pantoea* genera. In contrast, the analysis in Palmer et al. [@Palmer2017] identified *Mixta* outside of the other three genera. A separate study by Palmer et al. [@Palmer2018], Rezzonico et al. [@Rezzonico2016], and Brady et al. [@Brady2012] also ran analyses with the four MLSA genes as amino acid sequences. *Mixta* was found to be related closest to *Tatumella* by Rezzonico et al. [@Rezzonico2016] and *Pantoea* by Palmer et al. [@Palmer2018]. Brady et al. [@Brady2012] suggested that *Mixta* was a part of the *Pantoea* genus and not separate. Additionally, Palmer et al. [@Palmer2017] computed two ML trees; one used 1039 core amino acid sequences, and the other used 52 ribosomal MLSA nucleotide sequences. The 1039 core gene tree suggested *Pantoea* and *Tatumella* were sister taxa followed by two *Mixta* species. However, the ribosomal MLSA gene tree proposed *Pantoea* and *Tatumella* were sister taxa followed by *Erwinia* and then by *Mixta*. -->

<!-- What remains unclear is the relative contribution of disaparate evolutionary histories of *Mixta* genes, tree-building methods, and sequence evolution on the phylogenetic position of the genus. This paper aims to determine the cause behind the *Mixta* species' contentious phylogenetic results. Ten complete or partial genomes – two for each of *Mixta*, *Pantoea*, *Erwinia*, and *Tatumella* and two outgroups – were retrieved from a public database. Homologous genes were extracted from these genomes in both the nucleotide and amino acid sequences. Distance matrices were exported from the phylogenetic trees, and these were used to determine which species were most closely related to *Mixta* for each gene. Nucleotide and amino acid identity analyses were also conducted for each gene to compare the results between evolutionary history and sequence similarity. Phylogenetic trees were constructed for combinations of different gene sequences.  -->

<!-- # Materials and Methods -->
<!-- ## Genomes and Gene Sets -->
<!-- Ten publicly available full or partial genomes of the species' type strains were retrieved from the National Center for Biotechnology Information (NCBI). Two representative genomes were used for each of the four *Erwiniaceae* genera while *Enterobacter cloacae* (*Enterobacterales*) and *Pseudomonas syringae* (*Pseudomonaceae*) were used as outgroups (Table 1). The genomes were annotated with Prokka v1.14.1 [@Seemann2014] using default parameters, and homologous genes were extracted using GET_HOMOLOGUES software package [@Contreras-Moreira2013] with the bidirectional best-hit search algorithm using default parameters. GET_HOMOLOGUES extracted 954 homologous genes between the ten species. -->

<!-- ```{r Table1, echo=FALSE, message=FALSE, warning=FALSE, fig.cap=TRUE} -->
<!-- strain <- read.csv("C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Table1_Genomes.csv", stringsAsFactors = FALSE) -->

<!-- colnames(strain)[3] <- "GenBank Accession No." -->

<!-- kable(strain, caption = "Table 1. Genomes used in this study.") -->
<!-- ``` -->

<!-- In `R` [@RCore2019], two filter functions were written for the nucleotide sequence gene files. The first filter kept genes files that contained ten, single-copy sequences that were approximately the same length. The shortest gene length allowed was 90% the length of the longest sequence in the gene file. Of the 954 homologous gene sets, 38 multi-copy genes were excluded, and another 175 were excluded for not meeting the length requirements.  -->

<!-- A second filter was applied to these 175 genes to identify any genes wherein both representatives of each genera were approximately the same length. The filter split the genes into two groups: sequences that met the original 90% length cutoff and sequences that were between 80 and 90% the length of the longest sequence. If both representatives of a genus were in the same group, then the gene was added to the working dataset. If at least one of the sequences did not meet the 80% cutoff, then the gene was excluded. An additional 58 genes were added to the working dataset, resulting in 799 genes. The amino acid sequences for these genes were subsequently downloaded. -->

<!-- ## Gene Alignment -->
<!-- Gene alignments were done in `R` using a ClustalW algorithm. The fasta files were read into the session using the `readDNAStringSet()` function (package: Biostrings) [@Pages2020]. The function called to do the alignments was the `msaClustalW()` function (package: msa) [@Bodenhofer2015] using default parameters and 100 iterations. The aligned sequences were then converted into a comma-separated values (CSV)-compatible format using `msaConvert()` (package: msa)[@Bodenhofer2015] and written into a new fasta file using the `write.fasta()` function (package: seqinr) [@Charif2007] -->

<!-- ## Model Selection -->
<!-- The best model of sequence evolution was identified for each nucleotide and amino acid alignment using Model Selection as implemented in MEGA-CC [@Kumar2012]. Default parameters were used for model testing. The output was a CSV file for each gene. These CSV files were read into `R`, and the model for phylogenetic analysis was extracted based on the lowest Bayesian Information Criterion (BIC). -->

<!-- ## Phylogenetic Analysis -->
<!-- Phylogenetic analysis was performed in MEGA-CC using maximum likelihood with 500 bootstrap replications, the model, and rate patterns to create a phylogenetic tree for each gene. If a model required a Gamma distribution, the number of discrete Gamma categories was set to 5, which is the default. -->

<!-- Distance matrices were extracted from each of the phylogenetic trees and saved as Excel files. These files were then read into the `R` session. The genetic distances between *Mixta* and the other species were extracted for further analysis. Tables were created in `R Markdown` using the `kable()` function (package: knitr) [@Xie2020] and figures were created using the `ggplot()` function (package: ggplot2) [@Wickham2016]. -->

<!-- ## Nucleotide and Amino Acid Identity -->
<!-- Each of the non-*Mixta* sequences were aligned to one of the corresponding *Mixta* sequences in turn. Alignments were performed in R as described above, and the aligned sequences were then split into the individual sites using the `str_split()` function (package: stringr) [@Wickham2019]. A function was created using base `R` code to calculate the number of identical sites between each sequence and the *Mixta* homologues. Values that equal or are close to 100% are more identical to the *Mixta* sequences. -->

@@

## Phylogenetic Gene Trees for *secY* and *ttuB*
In the genetic distance analysis, only one (*ttuB*) and two (*secY* and *ttuB*) genes differed between the two *Mixta* species for the NTS and AAS, respectively. The phylogenetic trees for the *secY* gene show that the gene is well-conserved across the *Erwiniaceae* taxa (Figures 7A and 7B). When using the NTS, the *Mixta* species are grouped together and fall outside of the other three *Erwiniaceae* taxa (Figure 7A). When using the AAS, the *Mixta* species do not group together (Figure 7B). Additionally, *M. calida* has a slightly shorter branch length to the *Erwinia* species, whereas *M. gaviniae* has a somewhat shorter branch length to *P. septica* (Figure 7B). However, the nodes separating the *Mixta* species from the other taxa are not well-supported (Figures 7A and 7B). Unexpectedly, *P. septica*’s homologue of *secY* is most closely related to *E. cloacae* for the AAS (Figure 7B) whereas *E. cloacae* falls outside of the *Erwiniaceae* for the NTS (Figure 7A).

![Figure 7. Maximum-likelihood phylogenetic trees for members of the *Erwiniaceae* with *E. cloacae* and *P. syringiae* as outgroups. The trees were reconstructed from the nucleotide (**A**) and amino acid (**B**) sequences of the gene *secY*. HKY+G and LG+G models were used for the nucleotide and amino acid sequences, respectively. Both trees required a discrete Gamma distribution (5 categories) to model the evolutionary rate differences among sites. Branch support was inferred from 500 bootstrap replicates; the percentage of trees in which the associated taxa clustered together are adjacent to the branches. The tree is drawn to scale, with branch lengths measured in the number of substitutions per site. Evolutionary analyses were conducted in MEGAX.](C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Figure7_secY_Trees.png)

The phylogenetic trees for the *ttuB* gene show a clearer separation between the two *Mixta* homologues (Figures 8A and 8B). For both the NTS and AAS, *M. calida* was more closely related to *Tatumella* followed by *E. cloacae* and *M. gaviniae* was more closely related to *Pantoea* followed by *Erwinia* (Figures 8A and 8B). Unlike with *secY*, many of the nodes are well-supported (Figures 8A and 8B).

![Figure 8. Maximum-likelihood phylogenetic trees for members of the *Erwiniaceae* with *E. cloacae* and *P. syringiae* as outgroups. The trees were reconstructed from the nucleotide (**A**) and amino acid (**B**) sequences of the gene *ttuB*. GTR+G and LG+G models were used for the nucleotide and amino acid sequences, respectively. Both trees required a discrete Gamma distribution (5 categories) to model the evolutionary rate differences among sites. Branch support was inferred from 500 bootstrap replicates; the percentage of trees in which the associated taxa clustered together are adjacent to the branches. The tree is drawn to scale, with branch lengths measured in the number of substitutions per site. Evolutionary analyses were conducted in MEGAX.](C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Figure8_ttuB_Tree.png)

## Phylogenetic Trees from MLSA Genes
Phylogenetic analyses often use several concatenated essential gene sequences to infer relatedness, namely *atpD*, *gyrB*, *infB*, and *rpoB*. However, *atpD* and *gyrB* did not meet the length requirements during the filtering process. Therefore, genes encoding the other subunits of ATP synthase (*atp*), translation initiation factor IF (*inf*), and DNA-directed RNA polymerase (*rpo*) were concatenated to create phylogenetic trees. The resulting trees show that *Mixta* are more closely related to *Pantoea* followed by *Erwinia* (Figures 9A and 9B). With the NTS, a well-supported bifurcation occurs between *Mixta* and *Pantoea* (Figure 9A). However, this does not occur with the AAS (Figure 9B). Instead, a somewhat-supported bifurcation occurs first between *Pantoea* and *Erwinia* and then a little-supported bifurcation between these two and *Mixta* (Figure 9B).

![Figure 9. Maximum-likelihood phylogenetic trees for members of the *Erwiniaceae* with *E. cloacae* and *P. syringiae* as outgroups. The trees were reconstructed from the concantenated nucleotide (**A**) and amino acid (**B**) sequences of the genes *atpA*, *atpB*, *atpC*, *atpE*, *atpF*, *atpG*, *atpH*, *infA*, *infB*, *rpoA*, *rpoB*, *rpoC*, *rpoD*, and *rpoH*. GTR+G+I and LG+G+F models were used for the nucleotide and amino acid sequences, respectively. Both trees required a discrete Gamma distribution (5 categories) to model the evolutionary rate differences among sites. The nucleotide model allowed for some sites to be evolutionary invariable. Branch support was inferred from 500 bootstrap replicates; the percentage of trees in which the associated taxa clustered together are adjacent to the branches. The tree is drawn to scale, with branch lengths measured in the number of substitutions per site. Evolutionary analyses were conducted in MEGAX.](C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Figure9_Four_Tree.png)

The four analyses conducted in this study agreed on a closest relative for only four of these genes – *atpA*, *atpG*, *rpoB*, and *rpoC* – for both *M. calida* and *M. gaviniae* (Figures 10A and 10B). This closest relative was *P. septica* for these genes. The other genes were more closely related to *P. agglomerans*, *E. tasmaniensis*, *E. amylovora*, and *E. cloacae* and these varied amongst the analyses (Figures 10A and 10B). One only of the genes, *rpoD*, was said to be closely related to *E. cloacae* by one of the analyses; the NTS genetic distance analysis labelled both the *Mixta* species as being closely related to *E. cloacae* and the remaining three analyses labelled this gene as being closely related to *P. septica* (Figures 10A and 10B).

@@

## Phylogenetic Trees Showing Other Results

>>>>

\newpage
# Abstract
*Mixta* is a recently described genus within the bacterial family *Erwiniaceae* and is closely related to *Pantoea*, *Erwinia*, and *Tatumella*. Depending on the genes and method used for phylogenetic analysis, the genus with which *Mixta* shares the most recent ancestor differs. This study aimed to determine the cause behind these contentious results. Ten complete or partial genomes – two *Mixta* species, six species from other *Erwiniaceae* genera, and two outgroups – were retrieved from NCBI and annotated. Homologous genes were extracted yielding a dataset of 799 genes. The genes were aligned with the ClustalW algorithm, and MEGA-CC was used to calculate the most appropriate model, distance matrices, and phylogenetic trees for both nucleotide and amino acid sequences. Nucleotide and amino acid identity analyses were also done using the programming language `R`. *Pantoea* was the closest relative to the *Mixta* species in most analyses; however, results were not consistent. Some genes were also found to be more similar to other, non-*Pantoea* species. Diligence must be given to selecting genes for phylogenetic analysis and the method chosen to prevent any xenologous signal from distorting the actual relationships. Furthermore, future research should consider that different phylogenetic analyses may provide different results.

\newpage
# Introduction
The bacterial family *Erwiniaceae* includes diverse Gram-negative [@Winslow1917; @Prakash2015] organisms, including the genera *Mixta*, *Pantoea*, *Erwinia*, *Tatumella*, *Buchnera*, *Wigglesworthia*, *Phaseolibacter*, and *Kalamiella* [REF CRAIG]. *Mixta* is a relatively newer genus first identified by Palmer et al. in 2017 [@Palmer2017] and later described by the same researchers in 2018 [@Palmer2018]. Currently, there are only five *Mixta* species validly published [@LPSN]. *M. calida* and *M. gaviniae* (previously species of *Pantoea*) were initially found in infant formula and its production environment [@Popp2010]. Fritz et al. [@Fritz2014] demonstrated that some strains of *M. calida* were linked with post-surgical meningitis in humans. The other *Mixta* species include *M. theicola* which was isolated from black tea extract [@Kato2015], *M. intestinalis* which was isolated from a healthy human fecal sample [@Prakash2015], and *M. tenebrionis* which was isolated from the gut microbiota of a plastic-eating mealworm [@Xia2020]. The first two were previously named as *Pantoea* species.

*Mixta*'s position relative to other members of the *Erwiniaceae*, in particular to *Pantoea*, *Erwinia*, *Tatumella*, differs vastly between studies [@Prakash2015; @Palmer2017; @Palmer2018; @Adeolu2016; @Brady2011; @Rezzonico2016; @Brady2012; @Gueule2015]. Prakash et al. [@Prakash2015], Palmer et al. [@Palmer2017], Rezzonico et al. [@Rezzonico2016], and Gueule et al. [@Gueule2015] all used *gyrB*, *rpoB*, *atpD*, and *infB* nucleotide gene sequences in their phylogenetic analyses. Prakash et al. [@Prakash2015] conducted a neighbour-joining statistical method while the others used a maximum-likelihood (ML) method [@Palmer2017; @Rezzonico2016; @Gueule2015]. Three of the studies – Rezzonico et al. [@Rezzonico2016], Prakash et al. [@Prakash2015], and Gueule et al. [@Gueule2015] – placed the *Mixta* species within the *Pantoea* genera. In contrast, the analysis in Palmer et al. [@Palmer2017] identified *Mixta* outside of the other three genera. A separate study by Palmer et al. [@Palmer2018], Rezzonico et al. [@Rezzonico2016], and Brady et al. [@Brady2012] also ran analyses with the four MLSA genes as amino acid sequences. *Mixta* was found to be related closest to *Tatumella* by Rezzonico et al. [@Rezzonico2016] and *Pantoea* by Palmer et al. [@Palmer2018]. Brady et al. [@Brady2012] suggested that *Mixta* was a part of the *Pantoea* genus and not separate. Additionally, Palmer et al. [@Palmer2017] computed two ML trees; one used 1039 core amino acid sequences, and the other used 52 ribosomal MLSA nucleotide sequences. The 1039 core gene tree suggested *Pantoea* and *Tatumella* were sister taxa followed by two *Mixta* species. However, the ribosomal MLSA gene tree proposed *Pantoea* and *Tatumella* were sister taxa followed by *Erwinia* and then by *Mixta*.

What remains unclear is the relative contribution of disaparate evolutionary histories of *Mixta* genes, tree-building methods, and sequence evolution on the phylogenetic position of the genus. This paper aims to determine the cause behind the *Mixta* species' contentious phylogenetic results. Ten complete or partial genomes – two for each of *Mixta*, *Pantoea*, *Erwinia*, and *Tatumella* and two outgroups – were retrieved from a public database. Homologous genes were extracted from these genomes in both the nucleotide and amino acid sequences. Distance matrices were exported from the phylogenetic trees, and these were used to determine which species were most closely related to *Mixta* for each gene. Nucleotide and amino acid identity analyses were also conducted for each gene to compare the results between evolutionary history and sequence similarity. Phylogenetic trees were constructed for combinations of different gene sequences.

# Materials and Methods
## Genomes and Gene Sets
Ten publicly available full or partial genomes of the species' type strains were retrieved from the National Center for Biotechnology Information (NCBI). Two representative genomes were used for each of the four *Erwiniaceae* genera while *Enterobacter cloacae* (*Enterobacterales*) and *Pseudomonas syringae* (*Pseudomonaceae*) were used as outgroups (Table 1). The genomes were annotated with Prokka v1.14.1 [@Seemann2014] using default parameters, and homologous genes were extracted using GET_HOMOLOGUES software package [@Contreras-Moreira2013] with the bidirectional best-hit search algorithm using default parameters. GET_HOMOLOGUES extracted 954 homologous genes between the ten species.

```{r Table1, echo=FALSE, message=FALSE, warning=FALSE, fig.cap=TRUE}
strain <- read.csv("C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Table1_Genomes.csv", stringsAsFactors = FALSE)

colnames(strain)[3] <- "GenBank Accession No."

kable(strain, caption = "Table 1. Genomes used in this study.")
```

In `R` [@RCore2019], two filter functions were written for the nucleotide sequence gene files. The first filter kept genes files that contained ten, single-copy sequences that were approximately the same length. The shortest gene length allowed was 90% the length of the longest sequence in the gene file. Of the 954 homologous gene sets, 38 multi-copy genes were excluded, and another 175 were excluded for not meeting the length requirements.

A second filter was applied to these 175 genes to identify any genes wherein both representatives of each genera were approximately the same length. The filter split the genes into two groups: sequences that met the original 90% length cutoff and sequences that were between 80 and 90% the length of the longest sequence. If both representatives of a genus were in the same group, then the gene was added to the working dataset. If at least one of the sequences did not meet the 80% cutoff, then the gene was excluded. An additional 58 genes were added to the working dataset, resulting in 799 genes. The amino acid sequences for these genes were subsequently downloaded.

## Gene Alignment
Gene alignments were done in `R` using a ClustalW algorithm. The fasta files were read into the session using the `readDNAStringSet()` function (package: Biostrings) [@Pages2020]. The function called to do the alignments was the `msaClustalW()` function (package: msa) [@Bodenhofer2015] using default parameters and 100 iterations. The aligned sequences were then converted into a comma-separated values (CSV)-compatible format using `msaConvert()` (package: msa)[@Bodenhofer2015] and written into a new fasta file using the `write.fasta()` function (package: seqinr) [@Charif2007]

## Model Selection
The best model of sequence evolution was identified for each nucleotide and amino acid alignment using Model Selection as implemented in MEGA-CC [@Kumar2012]. Default parameters were used for model testing. The output was a CSV file for each gene. These CSV files were read into `R`, and the model for phylogenetic analysis was extracted based on the lowest Bayesian Information Criterion (BIC).

## Phylogenetic Analysis
Phylogenetic analysis was performed in MEGA-CC using maximum likelihood with 500 bootstrap replications, the model, and rate patterns to create a phylogenetic tree for each gene. If a model required a Gamma distribution, the number of discrete Gamma categories was set to 5, which is the default.

Distance matrices were extracted from each of the phylogenetic trees and saved as Excel files. These files were then read into the `R` session. The genetic distances between *Mixta* and the other species were extracted for further analysis. Tables were created in `R Markdown` using the `kable()` function (package: knitr) [@Xie2020] and figures were created using the `ggplot()` function (package: ggplot2) [@Wickham2016].

## Nucleotide and Amino Acid Identity
Each of the non-*Mixta* sequences were aligned to one of the corresponding *Mixta* sequences in turn. Alignments were performed in R as described above, and the aligned sequences were then split into the individual sites using the `str_split()` function (package: stringr) [@Wickham2019]. A function was created using base `R` code to calculate the number of identical sites between each sequence and the *Mixta* homologues. Values that equal or are close to 100% are more identical to the *Mixta* sequences.

@@

## Phylogenetic Gene Trees for *secY* and *ttuB*
In the genetic distance analysis, only one (*ttuB*) and two (*secY* and *ttuB*) genes differed between the two *Mixta* species for the NTS and AAS, respectively. The phylogenetic trees for the *secY* gene showed that the gene is well-conserved across the *Erwiniaceae* taxa (Figures 7A and 7B). When using the NTS, the *Mixta* species grouped together and fell outside of the other three *Erwiniaceae* taxa (Figure 7A). When using the AAS, the *Mixta* species did not group together (Figure 7B). Additionally, *M. calida* had a slightly shorter branch length to the *Erwinia* species, whereas *M. gaviniae* had a somewhat shorter branch length to *P. septica* (Figure 7B). However, the nodes separating the *Mixta* species from the other taxa were not well-supported (Figures 7A and 7B). Unexpectedly, *P. septica*’s homologue of *secY* was most closely related to *E. cloacae* for the AAS (Figure 7B) whereas *E. cloacae* fell outside of the *Erwiniaceae* for the NTS (Figure 7A).

![Figure 7. Maximum-likelihood phylogenetic trees for members of the *Erwiniaceae* with *E. cloacae* and *P. syringae* as outgroups. The trees were reconstructed from the nucleotide (**A**) and amino acid (**B**) sequences of the gene *secY*. HKY+G and LG+G models were used for the nucleotide and amino acid sequences, respectively. Both trees required a discrete Gamma distribution (5 categories) to model the evolutionary rate differences among sites. Branch support was inferred from 500 bootstrap replicates; the percentage of trees in which the associated taxa clustered together are adjacent to the branches. The tree is drawn to scale, with branch lengths measured in the number of substitutions per site. Evolutionary analyses were conducted in MEGAX.](C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Figure7_secY_Trees.png)

The phylogenetic trees for the *ttuB* gene showed a more apparent separation between the two *Mixta* homologues (Figures 8A and 8B). For both the NTS and AAS, *M. calida* was more closely related to *Tatumella* followed by *E. cloacae* and *M. gaviniae* was more closely related to *Pantoea* followed by *Erwinia* (Figures 8A and 8B). Unlike with *secY*, many of the nodes were well-supported (Figures 8A and 8B).

![Figure 8. Maximum-likelihood phylogenetic trees for members of the *Erwiniaceae* with *E. cloacae* and *P. syringae* as outgroups. The trees were reconstructed from the nucleotide (**A**) and amino acid (**B**) sequences of the gene *ttuB*. GTR+G and LG+G models were used for the nucleotide and amino acid sequences, respectively. Both trees required a discrete Gamma distribution (5 categories) to model the evolutionary rate differences among sites. Branch support was inferred from 500 bootstrap replicates; the percentage of trees in which the associated taxa clustered together are adjacent to the branches. The tree is drawn to scale, with branch lengths measured in the number of substitutions per site. Evolutionary analyses were conducted in MEGAX.](C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Figure8_ttuB_Tree.png)

## Phylogenetic Trees from MLSA Genes
Phylogenetic analyses often use several concatenated essential gene sequences to infer relatedness, namely *atpD*, *gyrB*, *infB*, and *rpoB*. However, *atpD* and *gyrB* did not meet the length requirements during the filtering process. Therefore, genes encoding the other subunits of ATP synthase (*atp*), translation initiation factor IF (*inf*), and DNA-directed RNA polymerase (*rpo*) were concatenated to create phylogenetic trees. The resulting trees showed that *Mixta* are more closely related to *Pantoea* followed by *Erwinia* (Figures 9A and 9B). With the NTS, a well-supported bifurcation occured between *Mixta* and *Pantoea* (Figure 9A). However, this did not happen with the AAS (Figure 9B). Instead, a somewhat-supported bifurcation occured first between *Pantoea* and *Erwinia* and then a weakly-supported bifurcation between these two and *Mixta* (Figure 9B).

![Figure 9. Maximum-likelihood phylogenetic trees for members of the *Erwiniaceae* with *E. cloacae* and *P. syringae* as outgroups. The trees were reconstructed from the concantenated nucleotide (**A**) and amino acid (**B**) sequences of the genes *atpA*, *atpB*, *atpC*, *atpE*, *atpF*, *atpG*, *atpH*, *infA*, *infB*, *rpoA*, *rpoB*, *rpoC*, *rpoD*, and *rpoH*. GTR+G+I and LG+G+F models were used for the nucleotide and amino acid sequences, respectively. Both trees required a discrete Gamma distribution (5 categories) to model the evolutionary rate differences among sites. The nucleotide model allowed for some sites to be evolutionary invariable. Branch support was inferred from 500 bootstrap replicates; the percentage of trees in which the associated taxa clustered together are adjacent to the branches. The tree is drawn to scale, with branch lengths measured in the number of substitutions per site. Evolutionary analyses were conducted in MEGAX.](C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Figure9_Four_Tree.png)

The four analyses conducted in this study agreed on the closest relative for only four of these genes – *atpA*, *atpG*, *rpoB*, and *rpoC* – for both *M. calida* and *M. gaviniae* (Figures 10A and 10B). This closest relative to these genes were *P. septica*. The other genes were more closely related to *P. agglomerans*, *E. tasmaniensis*, *E. amylovora*, and *E. cloacae* and these varied amongst the analyses (Figures 10A and 10B). One only of the genes, *rpoD*, was closely related to *E. cloacae* by one of the analyses; the NTS genetic distance analysis labelled both the *Mixta* species as being closely related to *E. cloacae* and the remaining three analyses labelled this gene as being closely related to *P. septica* (Figures 10A and 10B).

@@

## Phylogenetic Trees Showing Other Results
Several genes were found to be more closely related to taxa other than *P. septica* or *P. agglomerans*. Thus, gene sequences wherein the four analyses agreed upon each of the other taxa were concatenated to compute phylogenetic trees. There was a consensus for 32 genes that the closest relative was not either of the *Pantoea* species; 17 were more closely related to *E. tasmaniensis*, 8 to *E. amylovora*, 1 to *T. saanichensis*, and 6 to *E. cloacae*.

The genes most closely related to *E. tasmaniensis* were *fabG*, *folK*, *greB*, *hemE*, *hscA_2*, *ibaG*, *proA*, *rppH*, *sbp*, *tusA*, *tusE*, *uppS*, *ycfH*, *yffB*, and 3 hypothetical proteins. The phylogenetic trees resulting from the concatenated NTS (Figure 11A) and AAS (Figure 11B) showed strong support for the *Mixta* species being more closely related to *Erwinia*. Interestingly, *E. cloacae* fell outside of the *Erwiniaceae* for only the AAS (Figure 11B). However, the bifurcations between *E. cloacae* and the other taxa are not well-supported for either sequence type (Figures 11A and 11B).

![Figure 11. Maximum-likelihood phylogenetic trees for members of the *Erwiniaceae* with *E. cloacae* and *P. syringae* as outgroups. The trees were reconstructed from the concantenated nucleotide (**A**) and amino acid (**B**) sequences of the genes *fabG*, *folK*, *greB*, *hemE*, *hscA_2*, *ibaG*, *proA*, *rppH*, *sbp*, *tusA*, *tusE*, *uppS*, *ycfH*, *yffB*, and 3 hypothetical proteins. The four analyses in this study agreed on *E. tasmaniensis* as being the closest relative for these 17 genes. GTR+G+I and LG+G+F models were used for the nucleotide and amino acid sequences, respectively. Both trees required a discrete Gamma distribution (5 categories) to model the evolutionary rate differences among sites. The nucleotide model allowed for some sites to be evolutionary invariable. Branch support was inferred from 500 bootstrap replicates; the percentage of trees in which the associated taxa clustered together are adjacent to the branches. The tree is drawn to scale, with branch lengths measured in the number of substitutions per site. Evolutionary analyses were conducted in MEGAX.](C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Figure11_Etasmaniensis.png)

The eight genes most closely related to *E. amylovora* were *acuI*, *coaBC*, *dapB*, *grpE*, *mtnC*, *rlmJ*, *ssuE*, and a hypothetical protein. Similar to the results for *E. tasmaniensis*, the bifurcations between *Mixta* and *Erwinia* are well-supported for both the NTS (Figure 12A) and AAS (Figure 12B). Additionally, *E. cloacae*’s position relative to *Mixta* and *Erwinia* is well-supported for the NTS, but its position is ambiguous for AAS (Figures 12A and 12B).

![Figure 12. Maximum-likelihood phylogenetic trees for members of the *Erwiniaceae* with *E. cloacae* and *P. syringae* as outgroups. The trees were reconstructed from the concantenated nucleotide (**A**) and amino acid (**B**) sequences of the genes *acuI*, *coaBC*, *dapB*, *grpE*, *mtnC*, *rlmJ*, *ssuE*, and a hypothetical protein. The four analyses in this study agreed on *E. amylovora* as being the closest relative for these 8 genes. GTR+G+I and LG+G+F models were used for the nucleotide and amino acid sequences, respectively. Both trees required a discrete Gamma distribution (5 categories) to model the evolutionary rate differences among sites. The nucleotide model allowed for some sites to be evolutionary invariable. Branch support was inferred from 500 bootstrap replicates; the percentage of trees in which the associated taxa clustered together are adjacent to the branches. The tree is drawn to scale, with branch lengths measured in the number of substitutions per site. Evolutionary analyses were conducted in MEGAX.](C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Figure12_Eamylovora.png)

Only one gene, *tsf*, was most closely related to *T. saanichensis*. Nevertheless, this relationship is well-supported for both the NTS and AAS (Figures 13A and 13B). Unexpectedly, the two *Tatumella* species were not grouped for the AAS (Figure 13B).

![Figure 13. Maximum-likelihood phylogenetic trees for members of the *Erwiniaceae* with *E. cloacae* and *P. syringae* as outgroups. The trees were reconstructed from the concantenated nucleotide (**A**) and amino acid (**B**) sequences of the gene *tsf*. The four analyses in this study agreed on *T. saanichensis* as being the closest relative for this one gene. K2+G and LG+G models were used for the nucleotide and amino acid sequences, respectively. Both trees required a discrete Gamma distribution (5 categories) to model the evolutionary rate differences among sites. Branch support was inferred from 500 bootstrap replicates; the percentage of trees in which the associated taxa clustered together are adjacent to the branches. The tree is drawn to scale, with branch lengths measured in the number of substitutions per site. Evolutionary analyses were conducted in MEGAX.](C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Figure13_Tsaanichensis.png)

The six genes most closely related to the outgroup species *E. cloacae* were *bioD1*, *cueR*, *hemN_2*, *iscS*, *rpsM*, and *yqjC*. The relationship between *Mixta* and *E. cloacae* is well-supported in the phylogenetic trees computed from the concatenated NTS (Figure 14A) and AAS (Figure 14B). Both of the trees showed similar results as the *Tatumella* species were the next closest relative to *Mixta* with the same main difference being weaker branch support for the AAS (Figures 14A and 14B). Additionally, *Pantoea* and *Erwinia* fell into their own group (Figures 14A and 14B).
